Model for DES Type Ia Supernova Cosmology Analysis

### 2016-05-22 tex/toy.tex
recast the story in a cleaner way

The model has the following types of observables $O$ that
may be observed for each transient:
\begin{itemize}
\item $c$: Counts of the photometric measurements (light curves).
\item $s$: Counts of the spectroscopic measurements.
\item $z$: Redshift.
\item $T$: Transient type.
\end{itemize}
The measurements made of the observables are labeled  with a subscript ``o''.

We are interested in what the model predicts for the pdf of those observables.  The pdf is dependent on the population and we consider two such populations.  The first is
the underlying population
of all possible transients and observables accommodated by the model: the pdf for this
distribution is $p(O|M,M_I)$  and in this note unsubscripted $p$ and $P$'s are probabilities
for this popiulation.
The second is
the analysis sample that represents the population of transients and observables that end up being analyzed.
The analysis population differs from the underlying population in that the transients of the former 
must be discovered, pass photometric selection criteria, and for the SN~Ia sample be selected for spectroscopic
observation, have a successful type from the spectrum, and be typed as SN~Ia.
The $p$ and $P$ probabilities for the analysis population are subscripted with $S$.
For the SN~Ia sample the two are related by
\begin{align}
% \mathcal{L}_S(O_o | M, M_I) & = \frac{\epsilon(O_o, M, M_I) p(O_o|M, M_I)}{\int  \epsilon(O, M, M_I) p(O|M, M_I)dO},
\mathcal{L}_S(O_o | M, M_I) & = \frac{p_S(O_o|M, M_I)}{\int  p_S(O|M, M_I)dO},
\label{likelihood:eqn}
\end{align}
where
\begin{align}
p_S(O|M, M_I) &= p(O, disc=True, anal=True, sobs=True, type=True, T=SNIa |M, M_I)\\
&= p(O, disc=True, anal=True, sobs=True, type=True| T=SNIa, M, M_I)\\
& \times P(T=SNIa| M, M_I)\\
&= p(T, z, s, type=True| T=SNIa ,M, M_I)\\
& \times p(c, disc=True, anal=True, sobs=True| T=SNIa,  M, M_I)\\
& \times P(T=SNIa| M, M_I) \\
&= p(T, Œz, type=True|s,  T=SNIa, M, M_I) \\
& \times P(disc=True, anal=True, sobs=True| c, T=SNIa, M, M_I) \\
& \times p(s| T=SNIa, M, M_I) p(c| T=SNIa, M, M_I) P(T=SNIa| M, M_I).
\end{align}
We anticipate that the discovery, photometric selection, spectroscopic
targeting, and photometric measurements are independent of the typing success, and the
spectral, redshift and typing measurements.

As the sample-inclusion probabilities and the uncertainties in $O_o$ are expected to be uncorrelated between different transients
 the likelihood can be split into contributions
from each,
\begin{align*}
\mathcal{L}(O_o | M, M_I) & = \frac{\prod_{\alpha \in transients} p_S(O_o^\alpha|M, M_I)}{\prod_{\alpha \in transients} \int  p_S(O^\alpha|M, M_I)dO^\alpha}.
\end{align*}
Note that we work with counts:  correlated calibration uncertainties would prevent this splitting for fluxes.

A frequentist analysis can proceed with this likelihood.
Alternatively, the Bayesian posterior is specified by parameter priors as follows:
\begin{align*}
p(M, M_i | O_o) & \propto \mathcal{L}(O_o| M, M_I) p(M_I|M) p(M)\\
 & \propto \mathcal{L}(O_o| M, M_I) p(L| \tau, M) p(\tau|M) p(M).
\end{align*}
Once all the $p$'s and $P$'s are specified we are ready to analyze!

@@

\section{Probabilities: Minding Ones P's and p's}
Section~\ref{likelihood:sec} introduces the probabilities and probability distribution functions that describe the likelihood
and posterior for the model shown in Figure~\ref{pgm:fig}.
In this Section we present one possible model by specifying underlying the
probabilities; the assembly of other models should
resemble this example.

@@

The type of the object $\tau$ is drawn from the Bernoulli distribution 
\begin{equation}
\tau | p_{SNIa} \sim \mathcal{B}(1-p_{SNIa}),
\end{equation}
with the convention that $T=0$ corresponds to SN~Ia.
This is a term of $p_S(O|M, M_I)$

@@

\item  The luminosity distance comes from the physics-theory function
\begin{equation}
d_L = d_L(z; \Omega_M, w0),
\end{equation}
the flux for an observation at date $t$ is
\begin{equation}
f = 10^{-0.25 \left(M_{SNIa} - \alpha x_1 + \beta c \right)}\frac{1}{4\pi d_L^2} F_{SALT2}((t-t_0)/(1+z); x_0,  x_1, c, \lambda/(1+z)),
\end{equation}
and expected counts is
\begin{equation}
c_\star = 10^{Z/2.5}f.
\end{equation}
The distributions of counts are 
\begin{align*}
c | c_\star &\sim \mathcal{N}(c_\star, C_c)\\
s | c_\star &\sim \mathcal{N}(c_\star, \sigma_s),
\end{align*}
where $C_c$ and $C_s$ are the measurement covariances for the photometric and spectroscopic observations respectively.  These are terms of $p_S(O|M, M_I)$.

@@

\subsection{$p_S(O|M, M_I)$: Sample Selection Probability}
We have listed the probabilities for the underlying distribution.  We now describe the probabilities 
of the two terms that involve sample selection.

\begin{itemize}
\item Spectroscopic term: $p(T, z, type=True|s,  T=SNIa, M, M_I) $. 
This term can be reduced as
\begin{align*}
p(T, z, type=True|  s,  T=SNIa, M, M_I) &= p(T|  s,  T=SNIa, M, M_I) \\
&\times P(type=True| s,  T=SNIa, M, M_I)\\
&\times p(z|s,  T=SNIa, M, M_I),
 \end{align*}
with the (reasonable) approximation that the three measurements are
independent.

Condition §$type=True$: The probability of successful spectroscopic typing needs to be determined.
A simple possibility is that  it only depends on the signal-to-noise of the spectroscopic observation
$\epsilon_T(ston)$.  Then given a  spectroscopic noise of $\sigma_s$
\begin{equation}
P(type=True| s,  T=SNIa, M, M_I) = \epsilon_T(s/\sigma_s).
\end{equation}

Assembling all the terms gives
\begin{align*}
p(T, z, type=True|  s,  T=SNIa, M, M_I) &= \delta_{T, SNIa} \delta(z-z_{host}) \epsilon_T(s/\sigma_s) .
 \end{align*}

\item Photometric term: $P(disc=True, anal=True, sobs=True| c, T=SNIa, M, M_I)$.
\begin{align*}
P(disc=True, anal  =True, sobs=True| c, T=SNIa, M, M_I) & =P(anal=True| c, T=SNIa, M, M_I) \\
& \times P(sobs=True|  c, T=SNIa, M, M_I) \\
& \times P(disc=True| c, T=SNIa, M, M_I),
 \end{align*}
taking that the random processes that determine discovery, sample selection, and
spectroscopic observation are independent.

Condition $disc=True$:
The probability for discovery of a transient in a single image depends on the
signal-to-noise $\epsilon_D(ston)$.  The probability of discovery given all
the images is
\begin{equation}
P(disc=True | c, T=SNIa, M, M_I)
=  1-\prod_{\beta \in images} \left(1-\epsilon_D(c_\beta/N_\beta)\right).
\end{equation}

Condition $anal=True$:
As an illustrative example we write out the probability of a analysis-sample selection requiring two photometric points above
a signal-to-noise threshold $STON_T$: other sample selection cases are similar but with different combinatorics
\begin{align*}
P(anal=True  | c, T=SNIa, M, M_I)= &1 - \prod_{\beta \in images} (1-\mathcal{H}(c_\beta/N_\beta; STON_T))\\
&  - \sum_{\alpha \in images}\mathcal{H}(c_\alpha/N_\alpha; STON_T) \prod_{\beta \in \{images-\alpha\}} (1-\mathcal{H}(c_\beta/N_\beta; STON_T)),
 \end{align*}
 where $\mathcal{H}$ is the Heaviside step function.

Condition $sobs=True$:
The probability a transient has a spectroscopic observation.
Mat Smith and Chris D'Andrea should be able to provide this term.
Until we get this, lets make the simplifying ansatz that
\begin{align*}
P(sobs=True | c, T=SNIa, M, M_I)= & const.
 \end{align*}

>>>>

The model has the following types of observables $O$ that
may be observed for each transient:
\begin{itemize}
\item $c$: Counts of the photometric measurements (light curves).
\item $s$: Counts of the spectroscopic measurements.
\item $z$: Redshift.
\item $T$: Transient type.
\end{itemize}
The actual measurements of the observables are labeled  with a subscript ``o''.

We are interested in what the model predicts for the pdf of those observables
for the population of transients that enter the analysis.  
The probabilities for this ``analysis'' population are notated with the
subscript $S$, for example the likelihood is expressed as $\mathcal{L}_S(O| M, M_I)$.

The likelihood can be expressed in terms of probabilities of the observables
for all possible transients and observables accommodated by the model.
The probabilities for this ``underlying'' population are notated with no subscript,
so that the pdf of observables for this population is $p(O|M, M_I$).
The analysis population differs from the underlying population in that the transients of the former 
must, for the typed SN~Ia analysis, be discovered, pass photometric selection criteria, be selected for spectroscopic
observation, have a successful type from the spectrum, and be typed as SN~Ia.

The likelihood is proportional to the pdf of the underlying sample that
enter the analysis sample, as expressed by
\begin{align}
% \mathcal{L}_S(O_o | M, M_I) & = \frac{\epsilon(O_o, M, M_I) p(O_o|M, M_I)}{\int  \epsilon(O, M, M_I) p(O|M, M_I)dO},
\mathcal{L}_S(O | M, M_I) & \propto p(O, \text{in analysis}|M, M_I)\\
&  \propto \prod_{\alpha \in transients} p(O^\alpha, \text{in analysis}|M, M_I),
\end{align}
where the second line holds because the observations and stochastic decisions
for sample selection are expected to be independent from transient to transient.

For a single supernova labeled by $\alpha$,   the probability of the analysis population is related
to the underlying population by
\begin{align*}
p_S(O^\alpha|M, M_I) &\equiv p(O^\alpha, \text{in analysis}|M, M_I)\\
& = P(disc=True, anal=True, sobs=True, type=True, T^\alpha=SNIa|O^\alpha, M, M_I)p(O^\alpha|M, M_I)\\
& = P(disc=True, anal=True, sobs=True, type=True|T^\alpha=SNIa, O^\alpha, M, M_I)\\
& \times P(T^\alpha=SNIa| O^\alpha, M, M_I)p(O^\alpha|M, M_I)\\
&= P(disc=True|T^\alpha=SNIa,c^\alpha, M, M_I)
P(anal=True|T^\alpha=SNIa,c^\alpha, M, M_I) \\
&  \times P(sobs=True|T^\alpha=SNIa,c^\alpha, M, M_I)
 P(type=True|T^\alpha=SNIa, s^\alpha, M, M_I)\\
&  \times
P(T^\alpha=SNIa| M, M_I)p(c^\alpha|M, M_I)p(s^\alpha|M, M_I)p(z^\alpha|M, M_I).
\end{align*}
where we anticipate that the stochastic processes that enter each sample-selection step
are independent of each other, and that the measurement uncertainties are independent.
We find that $p_S$ is the pdf of the underlying
distribution $p(O^\alpha|M, M_I)$ multiplied by the efficiency of a realized target to
enter the analysis sample.

The likelihood can be expressed as
\begin{align}
\mathcal{L}_S(O_o | M, M_I) & = \frac{\prod_{\alpha \in transients}p_S(O_o^\alpha|M, M_I)}{\int  \prod_{\alpha \in transients} p_S(O^\alpha |M, M_I)dO}.
\label{likelihood:eqn}
\end{align}
A frequentist analysis can proceed with this likelihood.
Alternatively, the Bayesian posterior is specified by parameter priors as 
\begin{align*}
p(M, M_i | O_o) & \propto \mathcal{L}_S(O_o| M, M_I) p(M_I|M) p(M)\\
 & \propto \mathcal{L}_S(O_o| M, L, \tau) p(L| \tau, M) p(\tau|M) p(M).
\end{align*}
Once all the $p$'s and $P$'s are specified we are ready to analyze!

@@

\section{Probabilities: Minding Ones P's and p's}
Section~\ref{likelihood:sec} introduces the probabilities and probability distribution functions that describe the likelihood
and posterior for the model shown in Figure~\ref{pgm:fig}.
In this Section we present one possible model; the assembly of other models should
resemble this example.
We now drop the $\alpha$ label with the implicit understanding that the distributions
now described are per transient.

@@

The type of the object $\tau$ is drawn from the Bernoulli distribution 
\begin{equation}
\tau | p_{SNIa} \sim \mathcal{B}(1-p_{SNIa}),
\end{equation}
with the convention that $T=0$ corresponds to SN~Ia.

@@

\item  The luminosity distance comes from the physics-theory function
\begin{equation}
d_L = d_L(z; \Omega_M, w0),
\end{equation}
the flux for an observation at date $t$ is
\begin{equation}
f = 10^{-0.25 \left(M_{SNIa} - \alpha x_1 + \beta c \right)}\frac{1}{4\pi d_L^2} F_{SALT2}((t-t_0)/(1+z); x_0,  x_1, c, \lambda/(1+z)),
\end{equation}
and expected counts is
\begin{equation}
c_\star = 10^{Z/2.5}f.
\end{equation}
The distributions of counts are 
\begin{align*}
c | c_\star &\sim \mathcal{N}(c_\star, C_c)\\
s | c_\star &\sim \mathcal{N}(c_\star, \sigma_s),
\end{align*}
where $C_c$ and $C_s$ are the measurement covariances for the photometric and spectroscopic observations respectively.

@@

\subsection{Sample Selection Probability}
We have listed the terms that contribute to the probability that a target enters
the analysis sample.

\begin{itemize}
\item $P(T=SNIa| M, M_I) = \delta_{\tau, SNIa}$: The transient is
measured to be SN~Ia if it is SN~Ia.
\item $ P(type=True|T=SNIa, s, M, M_I)$: 
The probability of successful spectroscopic typing needs to be determined.
A simple possibility is that  it only depends on the signal-to-noise of the spectroscopic observation
$\epsilon_T(ston)$.  Then given a  spectroscopic noise of $\sigma_s$
\begin{equation}
P(type=True|  T=SNIa, s, M, M_I) = \epsilon_T(s/\sigma_s).
\end{equation}

\item $P(disc=True|T=SNIa,c, M, M_I)$:
The probability for discovery of a transient in a single image depends on the
signal-to-noise $\epsilon_D(ston)$.  The probability of discovery given all
the images is
\begin{equation}
P(disc=True | T=SNIa,c,  M, M_I)
=  1-\prod_{\beta \in images} \left(1-\epsilon_D(c_\beta/N_\beta)\right).
\end{equation}

\item $P(anal=True|T=SNIa,c, M, M_I)$:
As an illustrative example we write out the probability of a analysis-sample selection requiring two photometric points above
a signal-to-noise threshold $STON_T$: other sample selection cases are similar but with different combinatorics
\begin{align*}
P(anal=True  | T=SNIa, c,  M, M_I)= &1 - \prod_{\beta \in images} (1-\mathcal{H}(c_\beta/N_\beta; STON_T))\\
&  - \sum_{\alpha \in images}\mathcal{H}(c_\alpha/N_\alpha; STON_T) \prod_{\beta \in \{images-\alpha\}} (1-\mathcal{H}(c_\beta/N_\beta; STON_T)),
 \end{align*}
 where $\mathcal{H}$ is the Heaviside step function.

\item $P(sobs=True|T=SNIa,c, M, M_I)$:
The probability a transient has a spectroscopic observation.
Mat Smith and Chris D'Andrea should be able to provide this term.
Until we get this, lets make the simplifying ansatz that
\begin{align*}
P(sobs=True | T=SNIa, c, M, M_I)= & const.
 \end{align*}
